Canopy leaf area of a mature evergreen *Eucalyptus* woodland does not respond to elevated atmospheric [CO~2~] but tracks water availability

### 2015-08-06 manuscript.Rmd

\
We then estimated leaf area index (LAI) from $\tau_d$ ($L_\tau$), following Campbell and Norman [-@Campbell2000]. We assumed that leaves had a spherical leaf angle distribution, and leaves were randomly distributed in space. We recognize that these assumptions are very likely violated, but we account for this bias by calibrating the LAI estimate to leaf litter production as further detailed below. Diffuse transmittance of a canopy is a function of the LAI by integrating the Beer-Lambert law across the entire sky hemisphere, and accounting for differences in extinction coefficient with different viewing angles (because of leaf angle and longer path length through the canopy at low solar angles).

@@

where $\theta$ is the solar zenith angle, G($\theta$) is the foliage projection function in the direction of the beam [@Campbell2000]. Because we assumed a spherical leaf angle distribution, G($\theta$) = 0.5 [@Stenberg2006]. We solved numerically for LAI in Eq. 2, using a smooth integration over $\theta$, yielding $L_\tau$.

\
To overcome violations in the assumptions underlying Eq. 2, we calibrated $L_\tau$ against cumulative litter production during a 4-month drought period in 2013 (indicated in Fig. 1). During this period (14 July 2013 to 12 Nov 2013), no leaf growth occurred but LAI decreased steadily (and thus $\tau_d$ increased steadily, see Fig. 1). During the drought, only 42 mm precipitation was recorded, which was the second driest July-October period recorded at a nearby (<2km) weather station since 1881 (compared to a long-term average of 188 mm). A total of n = 11 days with sufficiently diffuse conditions to estimate $\tau_d$  were available during this period. Cumulative LAI leaf litter production ($P_L$, expressed in units of LAI) should match the observed decrease in LAI estimated from canopy transmittance. We calibrated Eq. 3,

@@

where *c* is a calibration parameter, $L_{\tau,t=0}$ is $L_\tau$ at the start of the drought period. We used linear regression without intercept to estimate *c* in Eq. 4 as `r f(calib)` (SE = `r f(tidy(caliblm)$std.error)`). It should be stressed that even after calibration with litter production, $L_\tau$ still includes stem and infrastructure area (see Terminology). However, this approximately constant area cancels out in the left-hand side of Eq. 3. We used a single calibration constant for all rings, because we did not find significant differences between the rings in terms of either litter production or change in $\tau_d$ during the 2013 drought period (see Fig. S1).

@@

\
For comparison between LAI from photos and PAR transmittance, we used linear regression. To test for effects of C~a~ treatment on $L_\tau$ and litter production, we used linear-mixed effects models with FACE ring as the random effect, and C~a~ treatment and date as factor variables.

\
To assess effects of eC~a~ on $L_\tau$, we accounted for pre-treatment differences in canopy structure, as follows. The average $L_\tau$ over the entire period was well correlated with basal area (BA) measured prior to the onset of the eC~a~ treatment across the FACE rings (Fig. 3). Because the EucFACE woodland is well established, the differences in BA are not due to C~a~ treatment effects, but instead represent natural variation within the woodland. We then calculated the deviation of actual $L_\tau$ from the fitted relationship with BA shown in Fig. 3, to yield a 'LAI anomaly'. This way, we effectively used BA as a covariate in the analysis to account for pre-treatment variation.

@@

\
The dynamics of $L_\tau$ were compared with litter fall, by estimating the absolute change in $L_\tau$ for every FACE ring, for every period when litter was collected. We did this by fitting a generalized additive model to $L_\tau$ for each ring separately (similar to that shown for $\tau_d$ in Fig. 1), and calculating the change in $L_\tau$ for each period ($\Delta L_\tau$). Litter production correlated significantly with $\Delta L_\tau$ when the latter was separated into periods where $\Delta L_\tau$ > 0 (net leaf growth) and $\Delta L_\tau$ <  0 (net leaf loss) (Fig. 4). During periods of net leaf loss, we found a significant correlation (with a slope of `r f(tidy(lmlitdown)$estimate[2])`, SE = `r f(tidy(lmlitdown)$std.error[2])`, n = `r glance(lmlitdown)$df.residual+2`), close to the 1:1 line, as expected since no leaf growth was observed during any period where LAI was decreasing. During periods of net leaf growth, we found a significant correlation (with a slope of `r f(tidy(lmlitup)$estimate[2])`, SE = `r f(tidy(lmlitup)$std.error[2])`, n = `r glance(lmlitup)$df.residual+2`) between $\Delta L_\tau$ and litter production (Fig. 5).

\
We estimated gross leaf production from measured litter production (P~L~) and temporal dynamics of $L_\tau$. For each litter collection period, the net change in $L_\tau$ was estimated using fitted generalized additive models to allow interpolation from a smoothed timeseries. Gross leaf production was then calculated as the sum of the net change in $L_\tau$ and the amount of leaf litter produced during that period. The temporal dynamics of litter and leaf production are shown in Fig. 5, and demonstrate the irregular temporal pattern in leaf production and P~L~. We found no significant difference in either litter or leaf production between C~a~ treatments at any time point (p > 0.2 for both, assessed with linear-mixed effects models).

\
Next, we related P~L~ to average $L_\tau$ (during the litter collection period) across the six FACE rings (Fig. 6a). If leaf lifespan is constant, P~L~ would be linearly related to $L_\tau$. Instead, we found that in eC~a~, P~L~ was lower at a given $L_\tau$ as compared to aC~a~ (Fig. 6a). We estimated mean leaf lifespan from Fig. 6a as the ratio of $L_\tau$ to annual total P~L~. We found a significant (p `r pval(LLtest$p.value)`) increase in leaf lifespan (LL) with elevated C~a~ (Fig. 6b), with LL = `r f(LLtest$estimate1)`yr in aC~a~ and `r f(LLtest$estimate2)`yr in eC~a~.

\
Finally, we quantified the relationship between timing of leaf flushing and environmental drivers. During the `r nummon` month study period, we observed four leaf flushing events in the field, with one before estimates of $L_\tau$ were available (Fig. 7). These events occurred in spring and summer, with no clear seasonal cycle (October 2012, January 2013, December 2013 and January 2015). The dynamics of $L_\tau$ from the sensor record matched these events observed in the field quite well (Fig. 7a), as estimated from a generalized additive model fitted to the average $L_\tau$ across the six FACE rings. From the fitted model, we estimated the numerical derivative of LAI (dL/dt), as well as its confidence interval. All four leaf flushing events corresponded to significant positive dL/dt. One brief significant positive dL/dt (June 2013) did not correspond to an observed leaf flushing event, but it was small in magnitude. It is also obvious that every leaf flush event occurred shortly after a significant rainfall event (Fig. 4 and 6), when the soil water content in the top 80 cm increased to ca. > 0.2 m^3^ m^-3^. Two periods of sustained decrease in LAI (red lines in Fig. 7b, during winter of 2013 and 2014) corresponded to periods of low soil water availability.

@@

\
We used the opportunity provided by the high quality timeseries of LAI to test a simple photography-based method suggested by Macfarlane et al. [-@Macfarlane2007]. The two methods compared quite well (R^2^ = `r f(glance(lmcanphot)$r.squared)`) (Fig. S2), suggesting both diffuse transmittance and the photography-based method can be reliably used in *Eucalyptus*-dominated woodlands to estimate LAI.

@@

@@

**Figure 2**. Average $L_\tau$ over the entire study period plotted against the basal area (sum of stem cross-sectional areas) of the FACE rings ($R^2$ = `r glance(balm)$r.squared`, LAI = `r tidy(balm)$estimate[1]` + `r tidy(balm)$estimate[2]` BA). This relationship was used to calculate the LAI anomaly in Fig. 3, as the difference between LAI predicted from this regression and the actual $L_\tau$ at any given time. Red symbols are elevated C~a~, blue symbols ambient C~a~.

@@

**Figure 3**.Time series of LAI after correcting for plot-to-plot differences in basal area (BA). At every time point, the LAI anomaly was calculated as $L_\tau$ minus LAI fitted from the relationship with BA shown in Fig. 2. Each symbol is a value for a FACE ring, but the smooth curve (fitted with a generalized additive model) takes into account the random plot effect. Note that the 95% confidence always overlap, denoting no significant treatment effect at $\alpha$ = 0.05, except for a brief period in late 2014. This difference was due to an insect attack that led to partial defoliation of the canopy in one FACE ring (see Fig. 1).

@@

**Figure 6**.  (a) Total litter production (converted from data in Fig. 5 to an annual estimate) for each of the FACE rings plotted against mean $L_\tau$ during the entire litter collection period. The ratio of mean $L_\tau$ to annual litter production gives the mean leaf life span. Grey lines indicate leaf lifespan isolines (in years). (b) Calculated leaf lifespan for the six FACE rings shows a significant difference between C~a~ treatments (t-test, p `r pval(LLtest$p.value)`).

@@

**Figure 7**. Dynamics of LAI in relation to environmental drivers. (a) Smoothed time series of $L_\tau$ averaged across all six FACE rings. (b) From panel (a), we calculated the daily change in LAI (dL/dt), including the 95% confidence interval for the derivative of the time series. Green indicates significant positive dL/dt, red indicates significant negative dL/dt (at $\alpha$ = 0.05). Four of the five significant positive increases corresponded to approximate dates of commencement of leaf flushing (horizontal lines) observed in the field (except the third brief significant increase). (c) Time course of soil water content measured over 0-80 cm soil depth interval, averaged across all sensors and all rings (blue line), and daily rainfall (black lines).  (d) Average daily air temperature with a fitted smoother.

>>>>

\
We then estimated leaf area index (LAI) from $\tau_d$ ($LAI_\tau$), following Campbell and Norman [-@Campbell2000]. We assumed that leaves had a spherical leaf angle distribution, and leaves were randomly distributed in space. We recognize that these assumptions are very likely violated, but we account for this bias by calibrating the LAI estimate to leaf litter production as further detailed below. Diffuse transmittance of a canopy is a function of the LAI by integrating the Beer-Lambert law across the entire sky hemisphere, and accounting for differences in extinction coefficient with different viewing angles (because of leaf angle and longer path length through the canopy at low solar angles).

@@

where $\theta$ is the solar zenith angle, G($\theta$) is the foliage projection function in the direction of the beam [@Campbell2000]. Because we assumed a spherical leaf angle distribution, G($\theta$) = 0.5 [@Stenberg2006]. We solved numerically for LAI in Eq. 2, using a smooth integration over $\theta$, yielding $LAI_\tau$.

\
To overcome violations in the assumptions underlying Eq. 2, we calibrated $LAI_\tau$ against cumulative litter production during a 4-month drought period in 2013 (indicated in Fig. 1). During this period (14 July 2013 to 12 Nov 2013), no leaf growth occurred but LAI decreased steadily (and thus $\tau_d$ increased steadily, see Fig. 1). During the drought, only 42 mm precipitation was recorded, which was the second driest July-October period recorded at a nearby (<2km) weather station since 1881 (compared to a long-term average of 188 mm). A total of n = 11 days with sufficiently diffuse conditions to estimate $\tau_d$  were available during this period. Cumulative LAI leaf litter production ($P_L$, expressed in units of LAI) should match the observed decrease in LAI estimated from canopy transmittance. We calibrated Eq. 3,

@@

where *c* is a calibration parameter, $L_{\tau,t=0}$ is $LAI_\tau$ at the start of the drought period. We used linear regression without intercept to estimate *c* in Eq. 4 as `r f(calib)` (SE = `r f(tidy(caliblm)$std.error)`). It should be stressed that even after calibration with litter production, $LAI_\tau$ still includes stem and infrastructure area (see Terminology). However, this approximately constant area cancels out in the left-hand side of Eq. 3. We used a single calibration constant for all rings, because we did not find significant differences between the rings in terms of either litter production or change in $\tau_d$ during the 2013 drought period (see Fig. S1).

\
We used the opportunity provided by the high quality timeseries of LAI to test a simple photography-based method suggested by Macfarlane et al. [-@Macfarlane2007]. The two methods compared quite well (R^2^ = `r f(glance(lmcanphot)$r.squared)`) (Notes S1, Fig. S3), suggesting both diffuse transmittance and the photography-based method can be reliably used in *Eucalyptus*-dominated woodlands to estimate LAI.

@@

\
For comparison between LAI from photos and PAR transmittance, we used linear regression. To test for effects of C~a~ treatment on $LAI_\tau$ and litter production, we used linear-mixed effects models with FACE ring as the random effect, and C~a~ treatment and date as factor variables.

\
To assess effects of eC~a~ on $LAI_\tau$, we accounted for pre-treatment differences in canopy structure, as follows. The average $LAI_\tau$ over the entire period was well correlated with basal area (BA) measured prior to the onset of the eC~a~ treatment across the FACE rings (Fig. 3). Because the EucFACE woodland is well established, the differences in BA are not due to C~a~ treatment effects, but instead represent natural variation within the woodland. We then calculated the deviation of actual $LAI_\tau$ from the fitted relationship with BA shown in Fig. 3, to yield a 'LAI anomaly'. This way, we effectively used BA as a covariate in the analysis to account for pre-treatment variation.

@@

\
The dynamics of $LAI_\tau$ were compared with litter fall, by estimating the absolute change in $LAI_\tau$ for every FACE ring, for every period when litter was collected. We did this by fitting a generalized additive model to $LAI_\tau$ for each ring separately (similar to that shown for $\tau_d$ in Fig. 1), and calculating the change in $LAI_\tau$ for each period ($\Delta L_\tau$). Litter production correlated significantly with $\Delta L_\tau$ when the latter was separated into periods where $\Delta L_\tau$ > 0 (net leaf growth) and $\Delta L_\tau$ <  0 (net leaf loss) (Fig. 4). During periods of net leaf loss, we found a significant correlation (with a slope of `r f(tidy(lmlitdown)$estimate[2])`, SE = `r f(tidy(lmlitdown)$std.error[2])`, n = `r glance(lmlitdown)$df.residual+2`), close to the 1:1 line, as expected since no leaf growth was observed during any period where LAI was decreasing. During periods of net leaf growth, we found a significant correlation (with a slope of `r f(tidy(lmlitup)$estimate[2])`, SE = `r f(tidy(lmlitup)$std.error[2])`, n = `r glance(lmlitup)$df.residual+2`) between $\Delta L_\tau$ and litter production (Fig. 5).

\
We estimated gross leaf production from measured litter production (P~L~) and temporal dynamics of $LAI_\tau$. For each litter collection period, the net change in $LAI_\tau$ was estimated using fitted generalized additive models to allow interpolation from a smoothed timeseries. Gross leaf production was then calculated as the sum of the net change in $LAI_\tau$ and the amount of leaf litter produced during that period. The temporal dynamics of litter and leaf production are shown in Fig. 5, and demonstrate the irregular temporal pattern in leaf production and P~L~. We found no significant difference in either litter or leaf production between C~a~ treatments at any time point (p > 0.2 for both, assessed with linear-mixed effects models).

\
Next, we related P~L~ to average $LAI_\tau$ (during the litter collection period) across the six FACE rings (Fig. 6a). If leaf lifespan is constant, P~L~ would be linearly related to $LAI_\tau$. Instead, we found that in eC~a~, P~L~ was lower at a given $LAI_\tau$ as compared to aC~a~ (Fig. 6a). We estimated mean leaf lifespan from Fig. 6a as the ratio of $LAI_\tau$ to annual total P~L~. We found a significant (p `r pval(LLtest$p.value)`) increase in leaf lifespan (LL) with elevated C~a~ (Fig. 6b), with LL = `r f(LLtest$estimate1)`yr in aC~a~ and `r f(LLtest$estimate2)`yr in eC~a~.

\
Finally, we quantified the relationship between timing of leaf flushing and environmental drivers. During the `r nummon` month study period, we observed four leaf flushing events in the field, with one before estimates of $LAI_\tau$ were available (Fig. 7). These events occurred in spring and summer, with no clear seasonal cycle (October 2012, January 2013, December 2013 and January 2015). The dynamics of $LAI_\tau$ from the sensor record matched these events observed in the field quite well (Fig. 7a), as estimated from a generalized additive model fitted to the average $LAI_\tau$ across the six FACE rings. From the fitted model, we estimated the numerical derivative of LAI (dL/dt), as well as its confidence interval. All four leaf flushing events corresponded to significant positive dL/dt. One brief significant positive dL/dt (June 2013) did not correspond to an observed leaf flushing event, but it was small in magnitude. It is also obvious that every leaf flush event occurred shortly after a significant rainfall event (Fig. 4 and 6), when the soil water content in the top 80 cm increased to ca. > 0.2 m^3^ m^-3^. Two periods of sustained decrease in LAI (red lines in Fig. 7b, during winter of 2013 and 2014) corresponded to periods of low soil water availability.

@@

@@

# Tables {.unnumbered}

**Table 1**. Results of a mixed-effects model of LAI with C~a~, Date and pre-treatment basal area (BA) as predictors. The model accounts for repeated measures by using FACE ring as the random effect. Sample size of LAI per ring was `r table(df$Ring)[1]`.

```{r results='asis', echo=FALSE}
lmer_LAI[,"p-value"] <- formatPval(lmer_LAI[,"p-value"])

kable(lmer_LAI)
```

@@

**Figure 2**. Average $LAI_\tau$ over the entire study period plotted against the basal area (sum of stem cross-sectional areas) of the FACE rings ($R^2$ = `r glance(balm)$r.squared`, LAI = `r tidy(balm)$estimate[1]` + `r tidy(balm)$estimate[2]` BA). This relationship was used to calculate the LAI anomaly in Fig. 3, as the difference between LAI predicted from this regression and the actual $LAI_\tau$ at any given time. Red symbols are elevated C~a~, blue symbols ambient C~a~.

@@

**Figure 3**.Time series of LAI after correcting for plot-to-plot differences in basal area (BA). At every time point, the LAI anomaly was calculated as $LAI_\tau$ minus LAI fitted from the relationship with BA shown in Fig. 2. Each symbol is a value for a FACE ring, but the smooth curve (fitted with a generalized additive model) takes into account the random plot effect. Note that the 95% confidence always overlap, denoting no significant treatment effect at $\alpha$ = 0.05, except for a brief period in late 2014. This difference was due to an insect attack that led to partial defoliation of the canopy in one FACE ring (see Fig. 1).

@@

**Figure 6**.  (a) Total litter production (converted from data in Fig. 5 to an annual estimate) for each of the FACE rings plotted against mean $LAI_\tau$ during the entire litter collection period. The ratio of mean $LAI_\tau$ to annual litter production gives the mean leaf life span. Grey lines indicate leaf lifespan isolines (in years). (b) Calculated leaf lifespan for the six FACE rings shows a significant difference between C~a~ treatments (t-test, p `r pval(LLtest$p.value)`).

@@

**Figure 7**. Dynamics of LAI in relation to environmental drivers. (a) Smoothed time series of $LAI_\tau$ averaged across all six FACE rings. (b) From panel (a), we calculated the daily change in LAI (dL/dt), including the 95% confidence interval for the derivative of the time series. Green indicates significant positive dL/dt, red indicates significant negative dL/dt (at $\alpha$ = 0.05). Four of the five significant positive increases corresponded to approximate dates of commencement of leaf flushing (horizontal lines) observed in the field (except the third brief significant increase). (c) Time course of soil water content measured over 0-80 cm soil depth interval, averaged across all sensors and all rings (blue line), and daily rainfall (black lines).  (d) Average daily air temperature with a fitted smoother.
